---
title: "Tidy Data Manipulation - AGAR 2018"
author: "Ellen Quillen"
date: "July 12, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Manipulation
Helpful data sheets for a variety of tidyverse packages including tidyr and ggplot2 are available here: https://www.rstudio.com/resources/cheatsheets/ 


#Tidy Data
Hadley Wickham's tretise on the logic of tidy data is here: http://vita.had.co.nz/papers/tidy-data.pdf. Tidy data has 

#Data Manipulation Examples in tidyr
These examples are heavily inspired by the tutorials available at https://github.com/rstudio/master-the-tidyverse/

```{r dplyr basics}
#install packages
#install.packages("tidyverse")

#load libraries
library(tidyverse)

#load data with base R as data.frame and readr as tibble
dir = "C:/Users/Ellen/Dropbox/AAAG/AGAR/AGAR 2018/R Viz/dplyr session/"

gene.counts.df<-read.table(file=paste0(dir,"example_gene_counts.txt"),header=TRUE)
gene.counts.df
head(gene.counts.df)

gene.counts<-read_tsv(file=paste0(dir,"example_gene_counts.txt"), col_names = TRUE, comment = "##", skip = 0, na = c("", "NA","."))
gene.counts
names(gene.counts)=c("Gene",names(gene.counts)[1:6])

gene.counts<-read_tsv(file=paste0(dir,"example_gene_counts.txt"), col_names = c("Gene","H0","H1","H2","C0","C1","C2"), na = c("", "NA","."),  skip = 1)
gene.counts

summary(gene.counts)
View(gene.counts)

gene.counts.df<-read.table(file=paste0(dir,"example_gene_counts.txt"),header=TRUE)
gene.counts<-rownames_to_column(gene.counts.df,"gene")
gene.counts<-as.tibble(gene.counts)
gene.counts

```

```{r tidytools}
#tidy verbs: filter, select, mutate, summarize, and arrange

sample.info<-read_tsv(file=paste0(dir,"example_sample_info.txt"))

sample.filtered<-filter(sample.info,Species=="human")
sample.filtered

sample.selected<-select(sample.filtered, -CellDensity)
sample.selected

sample.mutated<-mutate(sample.selected,scaledRNA=RNAConcentration/CellViability)
sample.mutated

sample.summary<-summarize(sample.mutated,avg=mean(scaledRNA))
sample.summary

#using pipe %>% 
myresult<-sample.info %>% 
  filter(Species=="human") %>% 
  select(-CellDensity) %>%
  mutate(scaledRNA=RNAConcentration/CellViability) %>%
  summarize(avg=mean(scaledRNA),max=max(scaledRNA),min=min(scaledRNA))

sample.arranged<-arrange(sample.mutated,scaledRNA)
sample.arranged
sample.arranged<-arrange(sample.mutated,desc(scaledRNA))
sample.arranged

```

```{r useful base functions}
#most base R functions can be called on tibbles
viability.t <- t.test(sample.info$CellViability ~ sample.info$Species)
viability.t
boxplot(sample.info$CellViability ~ sample.info$Species)
summary(lm(sample.info$RNAConcentration ~ sample.info$CellViability))
plot(sample.info$CellViability,sample.info$RNAConcentration)

sample.clean = filter(sample.info,CellViability > 0.50)
sample.clean
plot(sample.clean$CellViability,sample.clean$RNAConcentration)
abline(lm(sample.clean$RNAConcentration~sample.clean$CellViability))
```

```{r data wrangling examples}
#combining and selecting columns with select
sample.mapping<-read_tsv(file=paste0(dir,"example_read_mapping_summary.txt"))
sample.mapping
sample.highmap<-filter(sample.mapping,PropMapped>0.81)

left_join(sample.highmap, sample.clean, by="Samples")
right_join(sample.highmap, sample.clean, by="Samples")
inner_join(sample.highmap, sample.clean, by="Samples")
full_join(sample.highmap, sample.clean, by="Samples")

anti_join(sample.highmap, sample.clean, by="Samples")
anti_join(sample.clean, sample.highmap, by="Samples")

sample.highmap<-rename(sample.highmap, SampleName = Samples)
sample.data=full_join(sample.highmap, sample.clean, by=c("SampleName" = "Samples"))

select(sample.data,1:3)
select(sample.data, SampleName, CellDensity, CellViability)
vars<-c("CellDensity","CellViability", "SampleName")
select(sample.data,one_of(vars))
select(sample.data,contains("cell"))

#stringr allows regex matches, note TimePoint not included
select(sample.data,matches(".t."))

#creating new variables with mutate

#note this isn't a tidy tibble and so working "rowwise" isn't loved by dplyr (or R in general)
gene.counts<-mutate(gene.counts,ZeroCount = (H0 ==0) + (H1 ==0) + (H2 ==0) + (C0 ==0) + (C1 ==0) + (C2 ==0))
gene.counts

summarise_at(gene.counts,c("H0","H1","H2","C0","C1","C2"),var)

rv <- apply(gene.counts[,2:7], MARGIN=1, FUN=var)
length(rv)
rv<-as.tibble(rv)
gene.counts <- bind_cols(gene.counts,rv)
gene.counts
gene.counts<-rename(gene.counts,Variance=value)
gene.counts

#mutate and replace with log
gene.counts.log<-gene.counts %>% mutate_at(vars(matches(".[0-2]")), funs(log2(.+1)))
gene.counts.log

#subsetting tibbles with filtr and group_by
gene.counts.clean <- filter(gene.counts.log, ZeroCount > 2 & Variance > 5000)
gene.counts.clean

gene.counts.select <-filter(gene.counts,gene %in% c("ENSG00000001084", "ENSG00000001167", "ENSG00000000419"))
gene.counts.select

#group_by
sample.info %>% 
  group_by(Species) %>%
  summarise(mean.viability = mean(CellViability))

#saving data
write_delim(gene.counts.clean, file.path(dir, "gene_counts_clean.csv"),delim=",")

write_csv(gene.counts.clean, file.path(dir, "gene_counts_clean.csv"))

```


#Exercises
(1)	Read in human_v37_MT.gatk.called.filt_snps.vcf without the metadata
(2) Rename the first column
(3) Create a new tibble that only contains files that passed the filter
(4) Group by population and calculate average quality score (QUAL)
(5) Create separate tibbles for CEU, PUR, and YRI 
(6) Using gene.log from above, find minimum and maximum values for each gene and scale to 0-1 (subtract min and divide by max)


```{r functions}
#creating a function for frequently used sets of commands
clean_gene_counts <- function (x){
  myresult<-x %>% 
  filter(Species=="human") %>% 
  select(-CellDensity) %>%
  mutate(scaledRNA=RNAConcentration/CellViability) %>%
  summarize(avg=mean(scaledRNA),max=max(scaledRNA),min=min(scaledRNA))
}

clean_gene_counts(sample.info)
myresult

clean_return_counts <- function (x){
  myresult<-x %>% 
  filter(Species=="human") %>% 
  select(-CellDensity) %>%
  mutate(scaledRNA=RNAConcentration/CellViability) %>%
  summarize(avg=mean(scaledRNA),max=max(scaledRNA),min=min(scaledRNA))
  return(myresult)
}
clean_return_counts(sample.info)

```

#Exercises
(1) Write a function to remove loci QUAL < 2000 from a generic data set and apply it to each tibble from (5) above.

```{r iterations}
#lets make gene.counts into a tidy data set
dim(gene.counts.df)
t.gene.counts<-t(gene.counts.df)
t.gene.counts<-as.data.frame(t.gene.counts)
row.names(t.gene.counts)
t.gene.counts<-rownames_to_column(t.gene.counts,"gene")
t.gene.counts<-as.tibble(t.gene.counts)
head(names(t.gene.counts))
t.counts<-select(t.gene.counts,1:20)
t.counts

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

counts.rescale<-select(t.counts,gene)

for (i in 2:length(names(t.counts))) {
  value <- as.tibble(rescale01(t.counts[,i]))
  counts.rescale<-bind_cols(counts.rescale,value)
}

```

#Exercises
(1) Use a loop to find the min, max, and mean of each gene and put that in a new tibble. 

#Additional Resources
R for Data Science by Hadley Wickham & Garrett Grolemund (O'Reilly) in an outstanding reference and can be purchased in hardcopy or read online http://r4ds.had.co.nz/


